---
title: "Influential Factors of the Number of Days an Animal Spends at the Shelter"
author: "Hanfan Chen, Zhaohao Li, Zhenhao Qiao, Chao Wang, Rachael Watson"
output:
  pdf_document:
    number_sections: yes
fig_caption: yes
editor_options: 
  chunk_output_type: inline
---

```{r, include=FALSE}
knitr::opts_chunk$set(eval = T, echo = FALSE, comment = NA, message = FALSE, warning = FALSE)
```

# Introduction

Data on animals admitted to the Dallas animal shelter were collected over the course of a year, from October 2016 to September 2017. For each animal admitted to the shelter, the following information was recorded - the type of animal being admitted, the month and year it was admitted, the reason for the animal being admitted, the final outcome for the animal, whether the animal was micro-chipped, and the number of days the animal spent at the shelter.

This report will investigate which of these factors are influential in determining the number of days an animal spends at the shelter before its final outcome is decided.

```{r library}
library(ggplot2)
library(GGally)
library(tidyverse)
library(kableExtra)
library(epiDisplay)
library(pscl)
library(countreg)
library(sjPlot)
```

# Exploratory Data Analysis
The first five lines of the raw data:
```{r read the data}
data10 <- read.csv("dataset10.csv")

kable(head(data10, 5), caption = "Raw data") %>% 
  kable_styling(font_size = 10, latex_options = "hold_position")
```

Levels of each explanatory variable:
```{r}
for (i in 1:6) {
  cat(colnames(data10)[i],":\n")
  print(levels(as.factor(data10[,i])))
  cat("\n")
}
data10[,2] <- as.factor(data10[,2])
data10[,3] <- as.factor(data10[,3])
```
All the explanatory variables are categorical variables and each explanatory variable has multiple levels.

```{r hist of y, fig.cap = "\\label{fig:y_hist} Histogram of number of days spent at the shelter", fig.width = 10, fig.align = "center", fig.pos = "H"}
ggplot(data10, mapping = aes(x = time_at_shelter)) + 
  geom_histogram(bins = 25)
```
Figure \ref{fig:y_hist} displays the histogram of the response variable, which is the number of days spent at the shelter. The histogram shows evidence of the response variable being right-skewed and following a Poisson distribution.

```{r pair plots, fig.cap = "\\label{fig:pairs} Pair plots of the variables", fig.width = 10, fig.align = "center", fig.pos = "H"}
ggpairs(data10, lower="blank", axisLabels="none")
```
The explanatory variables are all categorical and their box plots are shown in Figure \ref{fig:pairs}. The median time at shelter appears to be low for all the explanatory variables, which is due to the median time at shelter being `r quantile(data10$time_at_shelter, prob=c(.5))`.


Since in Figure \ref{fig:y_hist} the response variable is right-skewed, a median of the response variable is calculated. The figures below display the median of each category of the different explanatory variables.
```{r bar plot 1, fig.cap = "\\label{fig:animal_bar} Bar plot of animal type vs median time at shelter", fig.width = 10, fig.align = "center", fig.pos = "H"}
animal_bar <- data10 %>% 
  group_by(animal_type) %>% 
  summarise(median = median(time_at_shelter))
ggplot(animal_bar, aes(x = animal_type, y = median, colour=animal_type, fill=animal_type)) +
  geom_col() +
  labs(x="Animal type", y="Time at Shelter")
```

```{r table of animal type summarise}
animal_categ <- data10 %>%
  group_by(animal_type) %>% 
  summarise(n=n(),Mean= mean(time_at_shelter), St.Dev = sd(time_at_shelter), Min=min(time_at_shelter), Q1 = quantile(time_at_shelter,0.25), Median=median(time_at_shelter), Q3 = quantile(time_at_shelter,0.75), Max=max(time_at_shelter)) 
kable(animal_categ, caption = "Summary statistics on the time at shelter by animal type") %>% 
  kable_styling(font_size = 10, latex_options = "hold_position")
```
From Figure \ref{fig:animal_bar}, the median value of time at shelter seems different for each category except cat and dog. This could be because the sample sizes for bird and wildlife are much smaller than those of dog and cat, so this result could be skewed.

```{r bar plot 2, fig.cap = "\\label{fig:month_bar} Bar plot of month vs median time at shelter", fig.width = 10, fig.align = "center", fig.pos = "H"}
month_bar <- data10 %>% 
  group_by(month) %>% 
  summarise(median = median(time_at_shelter))
ggplot(month_bar, aes(x = month, y = median, fill=month)) +
  geom_col() +
  labs(x="Month", y="Time at Shelter")
```

```{r table of month summarise}
month_categ <- data10 %>%
  group_by(month) %>% 
  summarise(n=n(),Mean= mean(time_at_shelter), St.Dev = sd(time_at_shelter), Min=min(time_at_shelter), Q1 = quantile(time_at_shelter,0.25), Median=median(time_at_shelter), Q3 = quantile(time_at_shelter,0.75), Max=max(time_at_shelter)) 
kable(month_categ, caption = "Summary statistics on the time at shelter by month") %>% 
  kable_styling(font_size = 10, latex_options = "hold_position")
```
From Figure \ref{fig:month_bar}, the median value of time at shelter is similar for each month. All the summary statistics are similar.

```{r bar plot 3, fig.cap = "\\label{fig:year_bar} Bar plot of year vs median time at shelter", fig.width = 10, fig.align = "center", fig.pos = "H"}
year_bar <- data10 %>% 
  group_by(year) %>% 
  summarise(median = median(time_at_shelter))
ggplot(year_bar, aes(x = year, y = median, fill=year)) +
  geom_col() +
  labs(x="Year", y="Time at Shelter")
```

```{r}
month_2016 <- data10 %>% 
  filter(year==2016)  
  
month_2017 <- data10 %>% 
  filter(year==2017) 
unique(as.character(unique(month_2016$month)) %in% as.character(unique(month_2017$month)))
```
There is no overlap between the months and years, since the data was recorded over the period of a year. According to Figure \ref{fig:year_bar}, there is no obvious difference between the two years and the relationship between the response variable and month variable is similar to the relationship between the response variable and the year variable. In fact, both variables represent the same information, namely when the animal was admitted. Therefore, the variable year is removed.

```{r}
data10 <- data10[,-3]
```

```{r bar plot 4, fig.cap = "\\label{fig:intake_bar} Bar plot of intake type vs median time at shelter", fig.width = 10, fig.align = "center", fig.pos = "H"}
intake_bar <- data10 %>% 
  group_by(intake_type) %>% 
  summarise(median = median(time_at_shelter))
ggplot(intake_bar, aes(x = intake_type, y = median, fill=intake_type)) +
  geom_col() +
  labs(x="Intake type", y="Time at Shelter")
```

```{r table of intake type summarise}
intake_categ <- data10 %>%
  group_by(intake_type) %>% 
  summarise(n=n(),Mean= mean(time_at_shelter), St.Dev = sd(time_at_shelter), Min=min(time_at_shelter), Q1 = quantile(time_at_shelter,0.25), Median=median(time_at_shelter), Q3 = quantile(time_at_shelter,0.75), Max=max(time_at_shelter)) 
kable(intake_categ, caption = "Summary statistics on the time at shelter by intake type") %>% 
  kable_styling(font_size = 10, latex_options = "hold_position")
```
From Figure \ref{fig:intake_bar}, an obvious difference is shown between each category.

```{r bar plot 5, fig.cap = "\\label{fig:outcome_bar} Bar plot of outcome type vs median time at shelter", fig.width = 10, fig.align = "center", fig.pos = "H"}
outcome_bar <- data10 %>% 
  group_by(outcome_type) %>% 
  summarise(median = median(time_at_shelter))
ggplot(outcome_bar, aes(x = outcome_type, y = median, fill=outcome_type)) +
  geom_col() +
  labs(x="Outcome type", y="Time at Shelter")
```

```{r table of outcome type summarise}
outcome_categ <- data10 %>%
  group_by(outcome_type) %>% 
  summarise(n=n(),Mean= mean(time_at_shelter), St.Dev = sd(time_at_shelter), Min=min(time_at_shelter), Q1 = quantile(time_at_shelter,0.25), Median=median(time_at_shelter), Q3 = quantile(time_at_shelter,0.75), Max=max(time_at_shelter)) 
kable(outcome_categ, caption = "Summary statistics on the time at shelter by outcome type") %>% 
  kable_styling(font_size = 10, latex_options = "hold_position")
```
Figure \ref{fig:outcome_bar} shows there is an obvious difference between each category. The sample size of DIED and FOSTER are small compared with the other categories.

```{r bar plot 6, fig.cap = "\\label{fig:chip_bar} Bar plot of chip status vs median time at shelter", fig.width = 10, fig.align = "center", fig.pos = "H"}
chip_bar <- data10 %>% 
  group_by(chip_status) %>% 
  summarise(median = median(time_at_shelter))
ggplot(chip_bar, aes(x = chip_status, y = median, fill=chip_status)) +
  geom_col() +
  labs(x="Chip status", y="Time at Shelter")
```

```{r table of chip status summarise}
chips_categ <- data10 %>%
  group_by(chip_status) %>% 
  summarise(n=n(),Mean= mean(time_at_shelter), St.Dev = sd(time_at_shelter), Min=min(time_at_shelter), Q1 = quantile(time_at_shelter,0.25), Median=median(time_at_shelter), Q3 = quantile(time_at_shelter,0.75), Max=max(time_at_shelter)) 
kable(chips_categ, caption = "Summary statistics on the time at shelter by chips status") %>% 
  kable_styling(font_size = 10, latex_options = "hold_position")
```
From Figure \ref{fig:outcome_bar}, some differences exist. The sample size of UNABLE TO SCAN is small compared with others.

# Formal Data Analysis——Fitting a Poisson model
```{r Poisson model}
poisson_model <- glm(time_at_shelter ~ ., data = data10, family = "poisson")
```

## Variable selection using AIC
```{r all subset variable selection using AIC for Poisson}
step(poisson_model, direction = "both")
```
Using AIC as a selection criteria, the model with the minimum AIC is selected and hence the best fit for the data is the saturated model.

## P-value and confidence intervals for the Poisson model
```{r Poisson model p-value}
summary(poisson_model)
```

```{r Poisson model CI plots, fig.cap = "\\label{fig:poisson_CIs} Confidence Intervals of the Poisson Model", fig.width = 10, fig.height = 7, fig.align = "center", fig.pos = "H"}
plot_model(poisson_model, show.values = T, transform = NULL, value.offset = 0.4, value.size = 5, dot.size = 3)
```
From Figure \ref{fig:poisson_CIs}, comparing with the baseline categories, according to the p-values. All the levels of the categorical variables intake type, outcome type and chip status are significant. Two levels are significant in the factor animal type and one is insignificant. Five out of eleven categories of month are significant and the others are not.

## Goodness of fit and overdispersion for the Poisson model
```{r Deviance and Chi-squared goodness-of-fit test}
poisgof(poisson_model)
```
Since the p-value is smaller than 0.05, the null hypothesis is rejected and the over-dispersion is significant.

A rootogram can be used to check the over-dispersion. It is easy to visualize whether the model is over-fitting or under-fitting the values using the zero line. If the bar is below the zero line then that value has been under-fitted. And if there is a space between the zero line and the bar then it has been over-fitted. For the model to be fitted correctly, the bar should sit as close to the zero line as possible.
```{r poisson rootogram,fig.cap = "\\label{fig:poisson_rootogram} Rootogram of the Poisson Model", fig.width = 10, fig.align = "center", fig.pos = "H"}
rootogram(poisson_model, max = max(data10$time_at_shelter))

mu <- predict(poisson_model, type = "response")
exp <- sum(dpois(x = 0, lambda = mu))
```
In Figure \ref{fig:poisson_rootogram}, the Poisson model is severely under-fitting zero counts. There were `r sum(data10$time_at_shelter < 1)` zero counts observed in the data set but the model only fitted `r round(exp)`. It is also over-fitting the lower positive counts and under-fitting the higher counts, suggesting there is over-dispersion due to excess zeroes in the model. Hence a hurdle model will be fitted to provide a better fit.

# Formal Data Analysis——Fitting a Hurdle model
## Fitting a Binomial-Poisson hurdle model
```{r Poisson Hurdle model}
hurdle_model <- hurdle(time_at_shelter ~ ., data = data10, zero.dist = "binomial", dist = "poisson")
summary(hurdle_model)
```

```{r Poisson hurdle rootogram,fig.cap = "\\label{fig:binomial_hurdle_rootogram} Rootogram of the Binomial Hurdle Model", fig.width = 10, fig.align = "center", fig.pos = "H"}
rootogram(hurdle_model, max = max(data10$time_at_shelter))
```
In Figure \ref{fig:binomial_hurdle_rootogram} counts 1,2 and 4 are being severely under-fitted, while 6-9 are being over-fitted. There is also under-fitting at the higher counts which suggests over-dispersion. Therefore, a negative binomial hurdle model shall be fitted to address this. 

## Fitting a Binomial-Negative binomial hurdle model
```{r fit negative binomial hurdle model}
hurdle_model_nb <- hurdle(time_at_shelter ~ ., data = data10, zero.dist = "binomial", dist = "negbin")
```

```{r negative binomial hurdle rootogram,fig.cap = "\\label{fig:negative_binomial_hurdle_rootogram} Rootogram of the Negative Binomial Hurdle Model", fig.width = 10, fig.align = "center", fig.pos = "H"}
rootogram(hurdle_model_nb, max = max(data10$time_at_shelter))
```
The AIC of the binomial hurdle model is `r format(round(AIC(hurdle_model)), scientific=FALSE)` and the AIC of the negative binomial hurdle model is `r round(AIC(hurdle_model_nb))`. From this, the negative binomial model shows a much better fit to the data. However, in Figure \ref{fig:negative_binomial_hurdle_rootogram} some values are still being under-fitted.

## Variable selection using AIC for negative binomial hurdle model
```{r all subset variable selection using AIC}
step(hurdle_model_nb, direction = "both")
```
Using AIC as a selection criteria, the model with the minimum AIC is selected and hence the best fit for the data is the model with animal type, chip status, intake type and outcome type as the explanatory variables.
```{r new negative binomial hurdle model}
hurdle_model_nb_2 <- hurdle(time_at_shelter ~ animal_type + intake_type + outcome_type + chip_status, data = data10, zero.dist = "binomial", dist = "negbin")
```

## P-value and confidence intervals for negative binomial hurdle model
```{r negative binomial hurdle model p-value}
summary(hurdle_model_nb_2)
```

```{r negative binomial hurdle model CI plots, fig.cap = "\\label{fig:hurdle_CIs} Confidence Intervals of the Negative Binomial Hurdle Model", fig.width = 10, fig.height = 8, fig.align = "center", fig.pos = "H"}
plot_model(hurdle_model_nb_2, show.values = T, transform = NULL, value.offset = 0.4, value.size = 5, dot.size = 3)
```
From Figure \ref{fig:hurdle_CIs}, comparing with the baseline categories, according to the p-values. 
In the Binomial model, all the levels of the categorical variables intake type and outcome type are significant, while all the levels of the categorical variables animal type and chip status are insignificant.
In the Truncated Poisson model, all the levels  of the categorical variable intake type are significant and all the levels of animal type are insignificant.

Since the variable animal type is not significant for the model, animal type is removed to fit a new model.
```{r new negative binomial hurdle model remove animal type}
final_hurdle_model_nb <- hurdle(time_at_shelter ~ intake_type + outcome_type + chip_status, data = data10, zero.dist = "binomial", dist = "negbin")
```
The AIC of the new model only increases by `r round(AIC(final_hurdle_model_nb) - AIC(hurdle_model_nb_2), 2)`, so the factor animal type is removed to make the model simpler.

```{r final negative binomial hurdle model CI plots, fig.cap = "\\label{fig:final_hurdle_CIs} Confidence Intervals of the Negative Binomial Hurdle Model", fig.width = 10, fig.height = 8, fig.align = "center", fig.pos = "H"}
plot_model(final_hurdle_model_nb, show.values = T, transform = NULL, value.offset = 0.4, value.size = 5, dot.size = 3)
```
From Figure \ref{fig:final_hurdle_CIs}, according to the p-value of each categorical variable, all the factors are influential.

## Goodness of fit for the negative binomial hurdle model
```{r final rootogram,fig.cap = "\\label{fig:final_model_rootogram} Rootogram of Negative Binomial Hurdle Model with reduced variables", fig.width = 10, fig.align = "center", fig.pos = "H"}
rootogram(final_hurdle_model_nb, style = "standing", max=max(data10$time_at_shelter))
```
The final model provides an adequate fit to the data. It has the lowest AIC of `r round(AIC(final_hurdle_model_nb), 2)` and as seen from Figure \ref{fig:final_model_rootogram}, the model, represented by the red line, fits most of the values of the count data well.
  
# Conclusions {#sec:Conc}
Due to the excess zeroes present in the data, the Poisson model is not a suitable fit to the data. The model which provides the best fit to the data is the negative binomial Hurdle model which includes intake type, outcome type and chip status as explanatory variables. Hence these factors are the most influential in determining the number of days an animal spends at the shelter before its final outcome is decided.
